---
title: "LEC 19 LMM"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(nlme)
library(ggplot2)
library (lattice)
```

## LMM

this file contains code for linear mixed effects models of the orthodontic example

* Note: we use the R pacakge nlme here. An alternative is the lme4 package. The lmer function in lme4 is more suitable for modeling multiple non-nested random effects

```{r read}
Orthodont <- read.table("orthodontic.dat", header = TRUE)
head(Orthodont)
```

## NLME

```{r model}
Orth.new <- groupedData(distance ~ age | child, data = as.data.frame(Orthodont)) # not necessary for lme

head(Orth.new)

OrthFem <- subset(Orth.new, male == 0)
OrthFem[1:5,]

plot(OrthFem) # default: ordered by max resp of each child

ggplot(Orth.new, aes(age, distance, group=child)) + geom_line()
```

## fit a random intercept model

```{r random_intercept}
LMM1 <- lme (distance ~ age, random = ~1 | child,  data = OrthFem, method='REML') 
summary (LMM1) # pay attention to: random effects, fixed effects, 
#
VarCorr(LMM1) # covariance estimates for random effects and variance for residuals
LMM1$sigma # std for residuals
vcov(LMM1) # covariance for fixed effects estimates (inv fisher info)
#
fixed.effects(LMM1) # fixed effects coeff 
random.effects(LMM1) # ordered random effects, BLUP (in this case, just b_i)
fitted(LMM1) # fixed+random for each subj in each visit
OrthFem$distance-fitted(LMM1) # residuals
LMM1$residuals
```

### For subject 1

```{r}
plot(OrthFem$age[1:4],OrthFem$distance[1:4],type='b',ylim=c(19,26),col=26) # original data for subj1
points(OrthFem$age[1:4],fixed.effects(LMM1)[1]+fixed.effects(LMM1)[2]*OrthFem$age[1:4], pch=4,col=26) # fixed effect model
lines(OrthFem$age[1:4],fitted(LMM1)[1:4], lty=5,type='b',pch=4,col=26) # mixed effect fits
random.effects(LMM1) # check random effects b_i for the first subj: -1.229
```

### For subject 2

```{r}
plot(OrthFem$age[5:8],OrthFem$distance[5:8],type='b',col=20) # original data for subj2
points(OrthFem$age[5:8],fixed.effects(LMM1)[1]+fixed.effects(LMM1)[2]*OrthFem$age[5:8], pch=4,col=20)
lines(OrthFem$age[5:8],fitted(LMM1)[5:8], lty=5,type='b',pch=4,col=20)
random.effects(LMM1) #check random effects b_i for the second subj: 0.340
```
### check equivalence to marginal model with compound symmetry correlation structure

```{r}
# check equivalence to marginal model with compound symmetry correlation structure
summary(gls(distance~age, OrthFem, correlation=corCompSymm(form = ~ 1 |child), method="REML"))
# check rho==sigma_b^2/(sigma_b^2+sigma^2)
```

## compare models (likelihood ratio test)
```{r}
LMM.1 <- lme (distance ~ age, random = ~ 1 | child,  data = OrthFem, method='ML') # do NOT use REML for likelihood ratio
LMM.2 <- lme (distance ~ 1, random = ~ 1 | child,  data = OrthFem, method='ML')
anova(LMM.2,LMM.1) 
```
## fit a random intercept and slope model

```{r}
LMM2 <- lme (distance ~ age, random = ~ 1+ age | child, data = OrthFem)
summary (LMM2) 
```

### Plot for subject 1 and 2
```{r}
#
plot(OrthFem$age[1:4],OrthFem$distance[1:4],type='b',ylim=c(19,26),col=26) # original data for child1
points(OrthFem$age[1:4],fixed.effects(LMM2)[1]+fixed.effects(LMM2)[2]*OrthFem$age[1:4], pch=4,col=26)
lines(OrthFem$age[1:4],fitted(LMM2)[1:4], lty=5,type='b',pch=4,col=26)
random.effects(LMM2) # check BLUP 
lines(OrthFem$age[5:8],OrthFem$distance[5:8],type='b',col=20) 
points(OrthFem$age[5:8],fixed.effects(LMM2)[1]+fixed.effects(LMM2)[2]*OrthFem$age[5:8], pch=4,col=20)
lines(OrthFem$age[5:8],fitted(LMM2)[5:8], lty=5,type='b',pch=4,col=20)
```

